---
title: "DATA 607 HW Week 6 - Tidy Data"
author: "Antonio J Bayquen"
date: "March 6, 2016"
output: html_document
---

**Load the relevant libraries**
```{r load-libs, eval=TRUE}
#make sure the Wickham libraries are loaded
library(tidyr)
library(dplyr)
library(ggplot2)
```

**Read CSV File**
```{r read-file, eval=TRUE}
#current directory
setwd("~/")
#read the csv file
flt_del <- read.csv("flight_delays.csv", header=T, stringsAsFactors = F)
#list the dataframe
flt_del
```

**Tidy the dataframe**
```{r tidy-data, eval=TRUE}

flt_del$X[2] = flt_del$X[1]
flt_del$X[4] = flt_del$X[3]

flt_tidy <- gather(flt_del,"City","Flights", 3:7)

flt_tidy
```

**Group data and summarize results of groupings **

```{r summ-data-1, eval=TRUE}
#Group data by City(Airport) and Airline and get the total flights (on time and delayed) per group
flt_sum1 <- flt_tidy %>% group_by(City,X) %>% summarise(flt_sum1 = sum(Flights))

#Group data by City(Airport) and get the total flights (on time and delayed) per group (City)
flt_sum2 <- flt_tidy %>% group_by(City) %>% summarise(flt_sum2 = sum(Flights))

#Group data by Airline and get the total flights (on time and delayed) per group Airline
flt_sum7 <- flt_tidy %>% group_by(X) %>% summarise(flt_sum7 = sum(Flights))

```


**Results show that about 75% of all the flights (in terms of the 5 airports listed) for American West are in Phoenix. For Alaska, most of its flights are in Seattle (about 55%)  - which is not as proportionately large as American West.**

```{r flts-per-city, eval=TRUE}

#Combine the 2 summaries using a join
flt_com1 <- inner_join(flt_sum1, flt_sum7, by="X")

#Relative Percentage of flight between Alaska and American West per city
flt_com1 %>% mutate(per_cit = round(flt_sum1/flt_sum7,2)) %>% select(City, X, per_cit)

#generate bar plot to show relative share for Alaska and American West per City
ggplot(flt_com1, aes(x = X, y=round((flt_sum1/flt_sum7),2), fill = City)) +  geom_bar(stat="identity",position="dodge") + xlab("Airlines") + ylab("Percentage of Flights per City") 

```


**Find the market share (in terms of number of flights) of Alaska and American West Relative to each other in the Aiports on the dataframe list.**
**While American West has a much stronger presence in southern cities such as Phoenix and San Diego, Alaska dominates American West in Seattle. It is a much closer competition for the 2 airlines in San Francisco and Los Angeles**

```{r mkt-shr, eval=TRUE}

#Combine the 2 summaries using a join
flt_com <- inner_join(flt_sum1, flt_sum2, by="City")

#Relative market share between Alaska and American West
flt_com %>% mutate(mkt_shr = round(flt_sum1/flt_sum2,2)) %>% select(City, X, mkt_shr)

#generate bar plot to show relative share for Alaska and American West per City
ggplot(flt_com, aes(x = X, y=round((flt_sum1/flt_sum2),2), fill = City)) +  geom_bar(stat="identity",position="dodge") + xlab("Airlines") + ylab("Market Shares (Relative to each other)") 

```


**Compare the delays/on time record in the five cities listed for the 2 airlines.
Alaska Airlines has a better on time arrival records when compared to American West for all the 5 cities (airports)**

```{r delays-1, eval=TRUE}
flt_com_del <- inner_join(flt_tidy, flt_sum1, by=c("City","X"))

#determine the percentage of flights that are on time and delayed per City per Airline
flt_com_del %>% mutate(del_per = round(Flights/flt_sum1,2)) %>% select(City, X, X.1, del_per)

#show percentage of delayed/on time flights in Los Angeles for the 2 airlines
flt_LA = filter(flt_com_del, City=="Los.Angeles")
ggplot(flt_LA, aes(x = X, y=round((Flights/flt_sum1),2), fill = X.1)) +  geom_bar(stat="identity", position="dodge") + xlab("Airlines") + ylab("Percentage Delayed/On Time(Los Angeles)")
                                                                                                                                               
#show percentage of delayed/on time flights in Phoenix for the 2 airlines
flt_Ph = filter(flt_com_del, City=="Phoenix")
ggplot(flt_Ph, aes(x = X, y=round((Flights/flt_sum1),2), fill = X.1)) +  geom_bar(stat="identity", position="dodge") + xlab("Airlines") + ylab("Percentage Delayed/On Time(Phoenix)")                                                                                                                                            #show percentage of delayed/on time flights in San Diego for the 2 airlines
flt_Ph = filter(flt_com_del, City=="San.Diego")
ggplot(flt_Ph, aes(x = X, y=round((Flights/flt_sum1),2), fill = X.1)) +  geom_bar(stat="identity", position="dodge") + xlab("Airlines") + ylab("Percentage Delayed/On Time(San Diego)")

#show percentage of delayed/on time flights in San Francisco for the 2 airlines
flt_Ph = filter(flt_com_del, City=="San.Francisco")
ggplot(flt_Ph, aes(x = X, y=round((Flights/flt_sum1),2), fill = X.1)) +  geom_bar(stat="identity", position="dodge") + xlab("Airlines") + ylab("Percentage Delayed/On Time(San Francisco)")

#show percentage of delayed/on time flights in Seattle for the 2 airlines
flt_Ph = filter(flt_com_del, City=="Seattle")
ggplot(flt_Ph, aes(x = X, y=round((Flights/flt_sum1),2), fill = X.1)) +  geom_bar(stat="identity", position="dodge") + xlab("Airlines") + ylab("Percentage Delayed/On Time(Seattle)")
```


**Although Alaska beat American West in all the 5 cities in the list, overall (for the all the flights arriving the 5 cities), American West has a better record. This can be attributed to the fact that the majority of American West's flights are in its hub in Phoenix (75%) where it has a excellent on time record (so does Alaska but it has a lot less flights to Phoenix). This disproportionately large percentage more than made up for a less stellar record in the other 4 airports. This is the reason why American West has the overall better record inspite of losing to Alaska in all the 5 airports in terms of on time record**

```{r total-delays, eval=TRUE}
#group flights by airline and on-time/delay status and sum the grouping
flt_sum3 <- flt_tidy %>% group_by(X,X.1) %>% summarise(flt_sum3 = sum(Flights))

#from the summary group abouve, group this by airline and sum the grouping
flt_sum4 <- flt_sum3 %>% group_by(X) %>% summarise(flt_sum4 = sum(flt_sum3))

#join the the 2 summary groupings
flt_todel <- inner_join(flt_sum3, flt_sum4, by=("X"))

#determine the overall delay/on time record for the 2 airlines
mutate(flt_todel, del_all = round(flt_sum3/flt_sum4,2)) %>% select(X, X.1,del_all)

#show the overall delay/on time record for the 2 airlines
ggplot(flt_todel, aes(x = X, y=round((flt_sum3/flt_sum4),2), fill = X.1)) +  geom_bar(stat="identity") + xlab("Airlines") + ylab("Overall Percentage of Delayed Flights") 
```
